---
title: "Homework 2"
author: "Mary Anna Kivenson"
date: "2/16/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(openintro)
```


Graded: 2.6, 2.8, 2.20, 2.30, 2.38, 2.44

## 2.6 Dice rolls
If you roll a pair of fair dice, what is the probability of

(a) getting a sum of 1?

P(sum of 1) = 0

(b) getting a sum of 5?

P(sum of 5) = 4 / 36 = 1 / 9

(c) getting a sum of 12?

P(sum of 12) = 1 / 36


## 2.8 Poverty and Language
The American Community Survey is an ongoing survey that
provides data every year to give communities the current information they need to plan investments
and services. The 2010 American Community Survey estimates that 14.6% of Americans live below
the poverty line, 20.7% speak a language other than English (foreign language) at home, and 4.2%
fall into both categories.59

(a) Are living below the poverty line and speaking a foreign language at home disjoint?

No, living below the poverty line and speaking a foreign language at home is not disjoint. As specified in the question, 4.2% of Americans fall into both categories, so clearly these two events can happen at the same time.

(b) Draw a Venn diagram summarizing the variables and their associated probabilities.

See venn diagram below.
```{r venn-diagram, echo=TRUE, message=FALSE, warning=FALSE}
library(VennDiagram)
grid.newpage()
venn.plot <- draw.pairwise.venn(area1      = 14.6,
                                area2      = 20.7,
                                cross.area = 4.2,
                                category   = c("Below the Povery Line (14.6%)", 
                                               "Speak a Second Language (20.7%)"),
                                fill       = c("blue", "red"),
                                cat.cex   = 1.4,
                                cat.pos   = c(0,0),
                                cat.dist  = c(-0.38,-0.45),
                                lwd       = 2, 
                                cex       = 2,
                                scaled     = TRUE)
```

(c) What percent of Americans live below the poverty line and only speak English at home?

P(below poverty line AND only speak english) = P(below poverty line) - P(below poverty line AND speak foreign language) = 14.6% - 4.2% = **10.4%**

(d) What percent of Americans live below the poverty line or speak a foreign language at home?

P(below povery line OR speak foreign language) = P(below poverty line) + P(speak foreign language) - P(below poverty line AND speak foreign language) = 14.6% + 20.7% - 4.2% = **31.1%**

(e) What percent of Americans live above the poverty line and only speak English at home?

P(above poverty line AND only speak english) = 1 - P(below povery line OR speak foreign language) = 1 - 31.1% = **68.9%**

(f) Is the event that someone lives below the poverty line independent of the event that the person speaks a foreign language at home?

P(below poverty line) x P(speak foreign language) = 14.6% x 20.7% = 3.02%, which does not equal P(below poverty line AND speak foreign language) = 4.2%. Therefore, the events are dependent. 


## 2.20 Assortative Mating
Assortative mating is a nonrandom mating pattern where individuals
with similar genotypes and/or phenotypes mate with one another more frequently than what would
be expected under a random mating pattern. Researchers studying this topic collected data on
eye colors of 204 Scandinavian men and their female partners. The table below summarizes the
results. For simplicity, we only include heterosexual relationships in this exercise.

(a) What is the probability that a randomly chosen male respondent or his partner has blue eyes?

P(male respondent or partner has blue eyes) = P(self blue) + P(partner blue) - P(self blue AND parter blue) = (108/204) + (114/204) - (78/204) = 144/204 = **36 / 51**

(b) What is the probability that a randomly chosen male respondent with blue eyes has a partner with blue eyes?

P(self blue AND parter blue) = 78 / 204 = **39 / 102**

(c) What is the probability that a randomly chosen male respondent with brown eyes has a partner with blue eyes? What about the probability of a randomly chosen male respondent with green eyes having a partner with blue eyes?

P(partner blue | self brown) = P(partner blue and self brown) / P(self brown) = (19 / 204) / (54 / 204)  = **19 / 54**  
P(partner blue | self green) = P(partner blue and self green) / P(self green) = (11 / 204) / (36 / 204) = **11 / 36**
   
(d) Does it appear that the eye colors of male respondents and their partners are independent? Explain your reasoning.

It does not appear that the eye colors of male respondents and their partners are independent. The tables below show actual and expected frequencies. Expected frequencies are calculated by taking the product of the respective row and column totals, divided by the grand total. We can see that the actual values differ from the expected results, however a chi square test may be necessary to make a more definite conclusion.

```{r exp-freq, echo=TRUE}
#actual
actual_table <- table(assortive.mating$self_male, assortive.mating$partner_female)
actual_table

#expected
contingency_table <- as.array(margin.table(actual_table,1)) %*% 
  t(as.array(margin.table(actual_table,2))) / margin.table(actual_table)
contingency_table
```


## 2.30 Books on a bookshelf
The table below shows the distribution of books on a bookcase
based on whether they are nonfiction or fiction and hardcover or paperback.

(a) Find the probability of drawing a hardcover book first then a paperback fiction book second when drawing without replacement.

P(hardcover book then paperback fiction) = (28/95) x (59/94) = 0.185

(b) Determine the probability of drawing a fiction book first and then a hardcover book second, when drawing without replacement.

P(fiction then hardcover) = P(hardcover fiction then hardcover) + P(paperback fiction then hardcover) = (13/95) x (27/94) + (59/95) x (28/94) = 0.2243

(c) Calculate the probability of the scenario in part (b), except this time complete the calculations under the scenario where the first book is placed back on the bookcase before randomly drawing the second book.

P(fiction then hardcover w/ replacement)  = (72/95) * (28/95) = 0.2234

(d) The final answers to parts (b) and (c) are very similar. Explain why this is the case.

The final answers in the last two answers are similar because the there are a lot of books on the shelf. As a result of the large population size, we get a small covariance between samples and the probabilities end up being similar. 


## 2.38 Baggage fees
An airline charges the following baggage fees: $25 for the first bag and
$35 for the second. Suppose 54% of passengers have no checked luggage, 34% have one piece of
checked luggage and 12% have two pieces. We suppose a negligible portion of people check more
than two bags.

(a) Build a probability model, compute the average revenue per passenger, and compute the
corresponding standard deviation.

Probability model:

```{r prob-model, echo=TRUE}
model = data.frame(matrix(vector(), 3, 3,
                dimnames=list(c(), c("outcomes", "probability", "cost"))),
                stringsAsFactors=F)
model$outcomes <- c(0,1,2)
model$probability <- c(0.54,0.34,0.12)
model$cost <- c(0,25,60)
model
```

Average revenue per passenger = 15.7:

```{r exp-value, echo=TRUE}
model$expected <- (model$probability * model$cost)
sum(model$expected)
```

Standard deviation = 19.95:

```{r st-dev, echo=TRUE}
model$variance <- (model$cost - sum(model$expected))^2 * model$probability
model
st_dev <- sqrt(sum(model$variance))
st_dev
```

(b) About how much revenue should the airline expect for a flight of 120 passengers? With what standard deviation? Note any assumptions you make and if you think they are justified.

The airline should expect a revenue of 120 x 15.70 = **$1,884** for 120 passengers. Assuming that the individual passengers in this problem are independent (which is justified because the amount of bags a passenger brings does not depend on other passengers), the standard deviation will be $218.5434.

```{r st-devb, echo=TRUE}

#expected mean
linearmean <- 120 * sum(model$expected)
linearmean 

#variance
linearst_dev <- sqrt(120 * sum(model$variance))
linearst_dev
```


## 2.44 Income and gender
The relative frequency table below displays the distribution of
annual total personal income (in 2009 inflation-adjusted dollars) for a representative sample of
96,420,486 Americans. These data come from the American Community Survey for 2005-2009.
This sample is comprised of 59% males and 41% females.

(a) Describe the distribution of total personal income.

The distribution of total personal income is normal and centered at 35,000 to 49,999 dollars.

(b) What is the probability that a randomly chosen US resident makes less than $50,000 per year?

P(less than $50,000) = 2.2% + 4.7% + 15.8% + 18.3% + 21.2% = 62.2%

(c) What is the probability that a randomly chosen US resident makes less than $50,000 per year and is female? Note any assumptions you make.

P(less than $50,000 AND female) = P(less than 50,000) AND P(female) = 62.2% x 41% = 25.5% assuming gender and income is independent.

(d) The same data source indicates that 71.8% of females make less than $50,000 per year. Use this value to determine whether or not the assumption you made in part (c) is valid.

The assumption I made in part c is not valid - the actual probability of being female and making less than 50,000 a year is significantly greater than my calculation.
